Transitivity of Subtyping for Intersection Types

### 2020-05-15 papers/LMCS-subtyping-intersections/response.txt

>> 1. lack of details for the motivating connection to Siek's broader
>>    project(s), alluded to in the Introduction but too briefly to
>>    assess the value of this work on that basis.

The revision includes a new section "Motivations from Verified
Compilation" to address this problem.

>> 2. failure, as also in the Laurent paper that seems to have been the
>>    starting point for this work, to consider the deep line of work on
>>    logical treatments of subtyping.
>>
>> ...  In contrast, Siek's system handles function types through some
>> very particular meta-level operations on type expressions and side
>> conditions to the function subtyping rule (->').  I worry that
>> these conditions will be brittle, and not work well in a setting
>> where compositionality at the type level is required.  Imagine, for
>> example, allowing type-level functions as in System F-omega.  Then
>> particular checks on the forms of type expressions will fail in the
>> presence of applications of type variables (i.e., abstracted type
>> expressions), since we will not have a concrete type expression to
>> apply these various checks and meta-level operations to.  So an
>> approach that is based just on syntactic conditions on type
>> expressions seems fragile and not likely to be extensible to richer
>> type systems.

The logical treatments of subtyping are indeed useful for extending to
richer type systems for programming languages. The author has made use
of them in the context of research on gradual typing.  However, the
motivation for the present work is rather different: constructing
filter models (denotational semantics) for programming languages. For
that purpose, there is no need for richer type systems such as System
F-omega; intersection types already provide what is needed to express
the runtime behavior of functional programs. The revision includes
changes to the introduction to differentiate the varied uses of
intersection type systems and to make it clear that the present
article's motivation comes from the realm of filter models, not type
systems for programming languages.

@@

I've changed the notation for \in and \subseteq to make it clear that
these notions are different from the usual ones and added some
explanation to provide an intuition for what they accomplish.

Reviewer 3
==========

>> The only question I had is specifically related to the recent adoption
>> of intersection types in Scala and in formal work on the underlying
>> DOT calculus. The need for transitivity elimination is also described
>> in the OOPSLA'16 paper "Type Soundness for Dependent Object Types",
>> and the apparent solution adopted there was to "push back" top-level
>> uses of the transitivity rule further up into derivations, based on a
>> lemma "A <: B /\ B <! C ==> A <! C" where "A <! C" is a subtyping
>> derivation that does not end in the transitivity rule but may use it
>> internally. While it is not possible to get rid of transitivity
>> entirely in DOT, it seems possible to apply the "pushback" process to
>> exhaustion in the BCD setting considered in this paper, and in fact
>> I'm wondering if this is roughly equivalent to the induction metric
>> "<<". Perhaps the author can give this some thought, and maybe a
>> discussion of this piece of related work would be a useful addition
>> (if only to strengthen the motivation).

That is an interesting connection. I think that this style of proof
would work using the new function rule (→′) but not with the old rules
(→) and (→∩). I've added a discussion of this connection to the
revision.

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
check

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes:
\begin{itemize}
  \item in type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
  \item in precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item in filter models for the denotational semantics of a variety
    of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{itemize}
The motivation for this article comes from the third use of
intersection types, in denotational semantics.  We are interested in
constructing filter models for functional programming languages in
support of mechanized compiler
verification. Section~\ref{sec:motivation} describes how this research
motivates the results in this article.

@@

The present article discusses how we were able to import the key
feature from Laurent's system, the rule for subtyping between function
types, into the BCD subtyping system, replacing the function $(\to)$,
distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
then prove transitivity without the sequence of six lemmas, instead
using one lemma that has been used previously to prove the inversion
principle for subtyping of function
types~\citep{Barendregt:2013aa}. The new subtyping system does not
technically satisfy the subformula property but it satisfies what we
call the \emph{subformula conjunction property}: every type occuring
in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
intersection of such subformulas.  This property is enough to ensure
that if $A$ and $B$ are self consistent, then so are all the types
that appear in the derivation of $A <: B$. As a result, the new
subtyping system is highly suitable for mechanizing filter models.

@@

There has been a recent flurry of interest in subtyping for
intersection types that has led to two algorithms that effectively
satisfy the subformula property, in addition to the one by
\citet{Laurent:2018aa} discussed above. These two algorithms were
developed for the purposes of type checking programming languages, and
not for the construction of filter models, so the goals are somewhat
different. In short, the algorithms introduce complications for the
sake of efficiency, which would be undesirable for use in a filter
model.

@@

\section{A New Subtyping Judgment}
\label{sec:new-subtyping}

Our new subtyping judgment relies on a few definitions that are given
in Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the
treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
(modulo associativity and commutativity) as if it were a set of types
$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
that $B$ \emph{contains} $A$ if $A \containedin B$.

@@

  \fbox{$\dom{A}, \cod{A}$}
  \begin{align*}
  \dom{A \to B} &= A \\
  \dom{A \cap B} &= \dom{A} \cap \dom {B} \\
  \\
  \cod{A \to B} &= B \\
  \cod{A \cap B} &= \cod{A} \cap \cod {B}
  \end{align*}

@@

\begin{figure}[tbp]
  \fbox{$A <: B$}
  \begin{gather*}
    \text{(refl$_\alpha$)} \; \inference{}{\alpha <: \alpha} \\[3ex]
    \text{(lb$_L$)} \; \inference{A <: C}{A \cap B <: C} \quad
    \text{(lb$_R$)} \; \inference{B <: C}{A \cap B <: C} \quad
    \text{(glb)} \; \inference{A <: C & A <: D}{A <: C \cap D} \\[3ex]
    (\to') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C \to D}
    \begin{array}{l}
      B \containedin A\\
      \neg\, \mathsf{top}(D) \\
      \neg \, \mathsf{topInCod}(B)
      \end{array}\\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
    (\TOP{\to}') \; \inference{}{A <: C \to D}\;\mathsf{top}(D)
  \end{gather*}
  \caption{The New Subtyping Judgment}
  \label{fig:new-subtyping}
\end{figure}

The new intersection subtyping judgment, $A <: B$, is defined in
Figure~\ref{fig:new-subtyping}. First, it does not include the (trans)
rule.  It also replaces the (refl) rule with reflexivity for atoms
(refl$_\alpha$). The most important rule is the one for function types
$(\to')$, which subsumes $(\to)$ and $({\to}{\cap})$ in BCD subtyping.
The $(\to')$ rule essentially turns the $\beta$-soundness property
into a subtyping rule~\citep{Barendregt:2013aa}. The $(\to')$ rule
says that a type $A$ is a subtype of a function type $C \to D$ if a
type contained in $A$, call it $B$, has domain and codomain that are
larger and smaller than $C$ and $D$, respectively. The use of
containment enables this rule to absorb uses of (incl$_L$) and
(incl$_R$) on the left.  The side conditions $\neg\;\topP{B}$ and
$\neg\;\topInCod{D}$ are needed because of the $(\TOP{\to}')$ rule,
which in turn is needed to preserve types under $\eta$-reduction.  In
the many systems that do not involve $\eta$-reduction, the
$(\TOP{\to}')$ rule can be omitted, as well as these side
conditions. The rules (lb$_L$) and (lb$_R$) adapt (incl$_L$) and
(incl$_R$) to a system without transitivity, and have appeared many
times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{\to}')$ rule
generalizes the $(\TOP{\to})$ rule, replacing the $\TOP$ on the left
with any type $A$, because for transitivity, any type is below
$\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in the
codomain on the right with any type $D$ that is equivalent to $\TOP$.

@@

The proof of transitivity relies on the following lemma, which is
traditionally needed to prove the inversion principle for function
types. However, it was not needed for our system because the rule
$(\to')$ is already quite close to the inversion principle.
The lemma states that if every function type $C \to D$ in A
factors $B$, then $\dom{A} \to \cod{A}$ also factors $B$.

@@

\begin{theorem}[Transitivity of $<:$]\label{thm:⊑-trans}
    If $A <: B$ and $B <: C$, then $A <: C$.
\end{theorem}
\begin{proof}
  The proof is by well-founded induction on the relation $\ll$.
  We proceed by cases on the last rule applied in the
  derivation of $B <: C$.
  \begin{description}
  \item[Case (refl$_\alpha$)] We have $B = C = \alpha$.  From the premise $A <:
    B$ we immediately conclude that $A <: \alpha$.
  \item[Case (lb$_L$)] So $B = B_1 \cap B_2$, $B_1 <: C$, and $A <: B_1 \cap
    B_2$.  We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
    \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
    hypothesis, noting that $\langle A, B_1, C \rangle \ll \langle A,
    B, C \rangle$
    %because $\size{B_1} < \size{B}$.
    because $\depth{B_1} \le \depth{B}$
    and $\size{B_1} < \size{B}$.
  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $B_2 <: C$, and $A <:
    B_1 \cap B_2$.  We have $A <: B_2$
    (Proposition~\ref{prop:subtyping} part \ref{prop:⊔⊑-inv}), so we
    conclude that $A <: C$ by the induction hypothesis, noting that
    $\langle A, B_2, C \rangle \ll \langle A, B, C \rangle$ because
    $\depth{B_2} \leq \depth{B}$ and $\size{B_2} < \size{B}$.

>>>>

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes:
\begin{enumerate}
  \item in type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
  \item in precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item in filter models for the denotational semantics of a variety
    of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{enumerate}
The motivation for this article comes from the third use of
intersection types, in denotational semantics.  We are interested in
constructing filter models for functional programming languages in
support of mechanized compiler
verification. Section~\ref{sec:motivation} describes how that research
motivates the results in this article.

@@

The present article discusses how we were able to import the key
feature from Laurent's system, the rule for subtyping between function
types, into the BCD subtyping system, replacing the function $(\to)$,
distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
then prove transitivity without the sequence of six lemmas, instead
using one lemma that has been used previously to prove the inversion
principle for subtyping of function
types~\citep{Barendregt:2013aa}. The new subtyping system does not
technically satisfy the subformula property but it satisfies what we
call the \emph{subformula conjunction property}: every type occuring
in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
intersection of such subformulas.  This property is enough to ensure
that if $A$ and $B$ are self consistent, then so are all the types
that appear in the derivation of $A <: B$.  Thus, the new subtyping
system is suitable for mechanizing filter models.

@@

There has been a recent flurry of interest in subtyping for
intersection types. We begin by discussing two algorithms that
effectively satisfy the subformula property. We then discuss issues
surrounding subtyping and transitivity in the Scala language.

@@

These two algorithms were developed for the purposes of type checking
programming languages, and not for the construction of filter models,
so the goals are somewhat different. In short, the algorithms
introduce complications for the sake of efficiency, which would be
undesirable for use in a filter model.

Research on the Scala programming language~\citep{Odersky:2004aa} has
led to the development of the Dependent Object Types (DOT) calculus,
which includes path-dependent types and intersection
types~\citep{Amin:2016aa,Amin:2017aa}.  The interplay between
path-dependent types, subtyping, and transitivity in DOT has proved a
challenge in the proof of type soundness. Nevertheless,
\citet{Rompf:2016aa} prove the inversion principle for function types
using a ``pushback'' lemma that reorganizes any subtyping derivation
such that the last rule to be applied is never transitivity. Applying
such a lemma exhaustively might provide an alternative route to
transitivity for the new subtyping system presented in this
article. Conversely, the new subtyping system might enable the
addition of distributivity of function and intersection types to
DOT. However, we suspect that the pushback approach would encounter
difficulties if applied to the original BCD subtyping rules.

\section{A New Subtyping System}
\label{sec:new-subtyping}

Our new subtyping system relies on a few definitions that are given in
Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the
treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
(modulo associativity and commutativity) as if it were a set of types
$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
that $B$ \emph{contains} $A$ if $A \containedin B$.

@@

  \fbox{$\dom{A}, \cod{A}$}
  \begin{align*}
  \dom{A \to B} &= A \\
  \dom{A \cap B} &= \dom{A} \,\cap\, \dom {B} \\
  \\
  \cod{A \to B} &= B \\
  \cod{A \cap B} &= \cod{A} \,\cap\, \cod {B}
  \end{align*}

@@

\begin{figure}[tbp]
  \fbox{$A <: B$}
  \begin{gather*}
    \text{(refl$_\alpha$)} \; \inference{}{\alpha <: \alpha} \\[3ex]
    \text{(lb$_L$)} \; \inference{A <: C}{A \cap B <: C} \quad
    \text{(lb$_R$)} \; \inference{B <: C}{A \cap B <: C} \quad
    \text{(glb)} \; \inference{A <: C & A <: D}{A <: C \cap D} \\[3ex]
    (\to') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C \to D}
    \begin{array}{l}
      B \containedin A\\
      \neg\, \mathsf{top}(D) \\
      \neg \, \mathsf{topInCod}(B)
      \end{array}\\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
    (\TOP{\to}') \; \inference{}{A <: C \to D}\;\mathsf{top}(D)
  \end{gather*}
  \caption{The New Subtyping System}
  \label{fig:new-subtyping}
\end{figure}

The new intersection subtyping system, with judgements of the form $A
<: B$, is defined in Figure~\ref{fig:new-subtyping}. First, it does
not include the (trans) rule.  It also replaces the (refl) rule with
reflexivity for atoms (refl$_\alpha$). The most important rule is the
one for function types $(\to')$, which subsumes $(\to)$ and
$({\to}{\cap})$ in BCD subtyping.  The $(\to')$ rule essentially turns
the $\beta$-soundness property into a subtyping
rule~\citep{Barendregt:2013aa}. The $(\to')$ rule says that a type $A$
is a subtype of a function type $C \to D$ if a type contained in $A$,
call it $B$, has domain and codomain that are larger and smaller than
$C$ and $D$, respectively. The use of containment enables this rule to
absorb uses of (incl$_L$) and (incl$_R$) on the left.  The side
conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
because of the $(\TOP{\to}')$ rule, which in turn is needed to
preserve types under $\eta$-reduction.  In the many systems that do
not involve $\eta$-reduction, the $(\TOP{\to}')$ rule can be omitted,
as well as these side conditions. The rules (lb$_L$) and (lb$_R$)
adapt (incl$_L$) and (incl$_R$) to a system without transitivity, and
have appeared many times in the literature~\citep{Bakel:1995aa}.  The
$(\TOP{\to}')$ rule generalizes the $(\TOP{\to})$ rule, replacing the
$\TOP$ on the left with any type $A$, because for transitivity, any
type is below $\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in
the codomain on the right with any type $D$ that is equivalent to
$\TOP$.

@@

Our proof of transitivity relies on the following lemma, which is
traditionally needed to prove the inversion principle for function
types. (In our system this lemma is not needed to prove the inversion
principle because the rule $(\to')$ is already quite close to the
inversion principle.)  The lemma states that if every function type $C
\to D$ in A factors $B$, then $\dom{A} \to \cod{A}$ also factors $B$.

@@

\begin{theorem}[Transitivity of $<:$]\label{thm:⊑-trans}
    If $A <: B$ and $B <: C$, then $A <: C$.
\end{theorem}
\begin{proof}
  The proof is by well-founded induction on the relation $\ll$.
  We proceed by cases on the last rule applied in the
  derivation of $B <: C$.
  \begin{description}
  \item[Case (refl$_\alpha$)] We have $B = C = \alpha$.  From the premise $A <:
    B$ we immediately conclude that $A <: \alpha$.
  \item[Case (lb$_L$)] So $B = B_1 \cap B_2$, $A <: B_1 \cap
    B_2$, and $B_1 <: C$.
    We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
    \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
    hypothesis, noting that $\langle A, B_1, C \rangle \ll \langle A,
    B, C \rangle$
    %because $\size{B_1} < \size{B}$.
    because $\depth{B_1} \le \depth{B}$
    and $\size{B_1} < \size{B}$.
  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $A <:
    B_1 \cap B_2$, and $B_2 <: C$.  We have $A <: B_2$
    (Proposition~\ref{prop:subtyping} part \ref{prop:⊔⊑-inv}), so we
    conclude that $A <: C$ by the induction hypothesis, noting that
    $\langle A, B_2, C \rangle \ll \langle A, B, C \rangle$ because
    $\depth{B_2} \leq \depth{B}$ and $\size{B_2} < \size{B}$.
